Differential cultural reproduction of skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2022-12-25 manuscript/manuscript.Rmd
minor revision including title change.

---
title: "Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment. We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. \\par \\textbf{Keywords:} Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021; @schillinger2017]. From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles. Of particular attention here are the experimental works conducted by Stout et al. [-@stout2014] focusing on inferring high knapping skill level and Garcia-Medrano et al. [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage. Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.

In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applied it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignoring other competing factors such as raw material constraints [@white1995], and c) being constrained by the basic fracture mechanics and design space of bifacial technology [@moore2011; @moore2016]. To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018]. The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of internal representation, where handaxes can emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]."

Until recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the concept of mental template in the study of handaxe morphology. Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning. They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms. This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al. [-@mcnabb2004], which highlighted the influence of individual agency that is complementary to the traditionally favored explanation of social learning. As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape. Regarding the critique of design space constraint, Moore and Perston's experiment [-@moore2016] suggested that bifaces can be manufactured through flake removals dictated by a random algorithm. However, Moore [-@moore2020: 656-657] also suggested that these random experiments cannot produce "attributes like the congruent symmetries of handaxes seen in the Late Acheulean." In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.

Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66]. This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989]. It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993]. As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem". This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action. The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by local practitioners. It is thus possible to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995]. When contextual information is less readily available as in the Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022]. In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of handaxe technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies. These two features have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].

@@

Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather reasonable variance ratio to avoid overfitting. Variable loadings (**Table** \@ref(tab:tab1)) indicate that the first principal component (PC1) captures overall cross-sectional thickness. It is positively correlated with all thickness measurements while negatively correlated with all other measurements. A higher PC1 value thus indicates a handaxe with thicker shape relative to width and length, and vice versa. The second principal component (PC2) tracks elongation and pointedness, as indicated by a positive covariance of maximum length and bottom width/thickness. As PC2 increases, a handaxe will be relatively longer and more convergent from the broad base to the tip. Thus, PC1 corresponds to cross-sectional thinning and PC2 to overall shape variation.

@@

We extracted the PC1 and PC2 values of individual handaxes and compared them between different groups, where the novice group was divided into three sub-groups based on their training stages as specified in the method section. As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.649$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot. Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is not statistically significant ($t=0.333$, $p>0.05$). An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.818$, $p>0.05$). These results essentially suggest that there is a significant difference between the pre-training group and post-training groups in both PC1 (skill level) and PC2 (mental template). However, the effects of training across different assessment periods on both dimensions are not significant. Regarding the delta weight of different groups, our analysis (**Figure** \@ref(fig:weight)) suggests that there is a significant difference between the pre-training group and Late training group, while all other pairwise group comparison results are insignificant. It can also be inferred that the expert group display a higher variability in terms of delta weight compared with novices.

@@

The negative correlation between the PC1 and PC2 values revealed a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form. Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes. In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface while not breaking the handaxe into half [-@callahan1979: 90]. As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed. It is possible that such constraints help to explain why our novice knappers on average produced more handaxes in similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove. Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers. This also provides an alternative explanation to the social transmission of form for the experimental convergence on pointed forms. In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.

>>>>

---
title: "Differential cultural reproduction of skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Rock Art Research Institute, School of Geography, Archaeology, and Environmental Studies, University of the Witwatersrand, Johannesburg, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment. We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural reproduction"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021; @schillinger2017]. From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles. Of particular attention here are the experimental works conducted by Stout et al. [-@stout2014] focusing on inferring knapping skill level and Garcia-Medrano et al. [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage. Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.

In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applied it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignoring other competing factors such as raw material constraints [@white1995], and c) being constrained by the basic fracture mechanics and design space of bifacial technology [@moore2011; @moore2016]. To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018]. The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of explicit internal representations of form, where the shape of handaxes can instead emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]."

Recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the concept of mental template in the study of handaxe morphology. Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations. From this perspective, social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning. They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms. This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al. [-@mcnabb2004], which highlighted the influence of individual agency that is complementary to the traditionally favored explanation of social learning. As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape. Regarding the critique of design space constraint, Moore and Perston's experiment [-@moore2016] suggested that bifaces can be manufactured through flake removals dictated by a random algorithm. However, Moore [-@moore2020: 656-657] also suggested that these random experiments cannot produce "attributes like the congruent symmetries of handaxes seen in the Late Acheulean." In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.

Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66]. This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989]. It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993]. As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem". This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action. The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by local practitioners. It is thus possible to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995]. When contextual information is less readily available as in the Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022]. In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of handaxe technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019; @whittaker2004: 180-182] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies. These two features have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].

@@

Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather reasonable variance ratio to avoid overfitting. Variable loadings (**Table** \@ref(tab:tab1)) indicate that the first principal component (PC1) captures relative cross-sectional thickness ("refinement"). It is positively correlated with all thickness measurements while negatively correlated with all other measurements. A higher PC1 value thus indicates a handaxe that is thicker relative to width and length, and vice versa. The second principal component (PC2) tracks elongation and pointedness, as indicated by a positive covariance of maximum length and bottom width/thickness. As PC2 increases, a handaxe will be relatively longer and more convergent from the broad base to the tip. Thus, PC1 corresponds to cross-sectional thinning and PC2 to a narrowing of the tip relative to length and base dimensions.

@@

We extracted the PC1 and PC2 values of individual handaxes and compared them between different groups, where the novice group was divided into three sub-groups based on their training stages as specified in the method section. As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.649$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot. Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is not statistically significant ($t=0.333$, $p>0.05$). An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.818$, $p>0.05$). These results essentially suggest that there is a significant difference between the pre-training group and post-training groups in both PC1 (thinning) and PC2 (pointedness). However, the effects of training across different assessment periods on both dimensions are not significant. Regarding the delta weight of different groups, our analysis (**Figure** \@ref(fig:weight)) suggests that there is a significant difference between the pre-training group and Late training group, while all other pairwise group comparison results are insignificant. It can also be inferred that the expert group display a higher variability in terms of delta weight compared with novices.

@@

The negative correlation between the PC1 and PC2 values revealed a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form. Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes. In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface while not breaking the handaxe into half [-@callahan1979: 90]. As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed, echoing the high technological difficulty of making large yet thin bifacial points as perceived by American hobbyist flintknappers [@whittaker2004: 180-182]. It is possible that such constraints help to explain why our novice knappers on average produced more handaxes in similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove. Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers. This also provides an alternative explanation to the social transmission of form for the experimental convergence on pointed forms. In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.
